---
title: "EDA-challenge.Rmd"
output:
  pdf_document: default
  html_document: default
date: "2024-02-07"
---
```{r}
library(curl)
f <- curl("https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/data-wrangling.csv") 
d <-  read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE) #Loading "data-wrangling.csv" as a tabular data structure named d
```

```{r}
names(d) #Looking over the variables it contains
```
### **Step 1**

```{r}
library(tidyverse)
```

**1. Here, I'm creating a new variable named BSD (body size dimorphism) which is the ratio of average male to female body mass.**
```{r}
d$BSD <- d$Body_mass_male_mean/d$Body_mass_female_mean
```
**2. Here, I'm creating a new variable named sex_ratio, which is the ratio of the number of adult females to adult males in a typical group.**
```{r}
d$sex_ratio <- d$AdultFemale/d$AdultMales
```
**3. Here, I'm creating a new variable named DI (for “defensibility index”), which is the ratio of day range length to the diameter of the home range.**
```{r}
d$DI <- d$DayLength_km/(2*(sqrt((d$HomeRange_km2)/pi)))
```
**4. Here, I'm plotting the relationship between log(day range length) and log(time spent moving) for these primate species overall.**
```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
p <- ggplot(data = d, aes( #Building plot object
  x = log(DayLength_km), #I am log-transforming the variables to reduce the skew of the distribution
  y = log(Move),
)) 
p <- p + 
  xlab("log(Day Range Length)") +
  ylab("log(Time Spent Moving)") + #Specifying axis labels
  geom_point(na.rm = TRUE) + #Creating my scatterplot
  geom_smooth(method = "lm", se=FALSE, na.rm = TRUE) #Adding a linear regression model
p #Plotting the object
```

Looking at these primate species overall, there does not appear to be a strong correlation between time spent moving and day range length. 

Now, I'm plotting the relationship between log(day range length) and log(time spent moving) for these these primate species by family.
```{r}
p <- ggplot(data = d, aes(
  x = log(DayLength_km), 
  y = log(Move),
  color = factor(Family) #Coloring points by family
)) 
p <- p + 
  xlab("log(Day Range Length)") +
  ylab("log(Time Spent Moving)") + 
  geom_point(na.rm = TRUE) + 
  facet_wrap(~Family, ncol = 4) + #I am wrapping the data by family (14 total) and arranging these subset plots into 4 columns 
  theme(legend.position = "none") + #Because of the above, I do not need a legend to clarify the color assignments by family
  geom_smooth(method = "lm", fullrange = FALSE, se=FALSE, na.rm = TRUE)#I'm going to add a linear regression model to each famiy plot
p
```

Within Atelidae, Cebidae, and Hylobatidae, it does appear that there could be a trend for species that spend more time moving to travel farther.

**5. Here, I'm plotting the relationship between log(day range length) and log(mean group size) for these primate species overall.**
```{r}
p <- ggplot(data = d, aes(
  x = log(MeanGroupSize),
  y = log(DayLength_km),
)) 
p <- p + 
  xlab("log(Mean Group Size)") +
  ylab("log(Day Range Length)") + 
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm", se=FALSE, na.rm = TRUE)
p
```

Looking at these primate species overall, there does not appear to be a strong correlation between mean group size and day range length. 

Now, I'm plotting the relationship between log(day range length) and log(mean group size) by family.
```{r}
p <- ggplot(data = d, aes(
  x = log(MeanGroupSize), 
  y = log(DayLength_km),
  color = factor(Family) 
)) 
p <- p + 
  xlab("log(Mean Group Size)") +
  ylab("log(Day Range Length)") + 
  geom_point(na.rm = TRUE) + 
  facet_wrap(~Family, ncol = 4) +
  theme(legend.position = "none") +
  geom_smooth(method = "lm", fullrange = FALSE, se=FALSE, na.rm = TRUE)
p
```

Particularly within Atelidae, Cebidae, Cercopithecidae, Hominidae, and Pitheciidae, it does appear that there could be a trend for species that live in larger groups to travel farther.

**6. Here, I'm plotting the relationship between log(body size dimorphism) and log(canine size dimorphism) for these primate species overall.**
```{r}
p <- ggplot(data = d, aes(
  x = log(BSD),
  y = log(Canine_Dimorphism),
)) 
p <- p + 
  xlab("log(Body Size Dimorphism)") +
  ylab("log(Canine Size Dimorphism)") + 
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm", se=FALSE, na.rm = TRUE)
p
```

Looking at these primate species overall, it does appear that there could be a trend for species with greater body size dimorphism to also show greater canine size dimorphism.

Now, I'm plotting the relationship between log(body size dimorphism) and log(canine size dimorphism) by family.
```{r}
p <- ggplot(data = d, aes(
  x = log(BSD), 
  y = log(Canine_Dimorphism),
  color = factor(Family) 
)) 
p <- p + 
  xlab("log(Body Size Dimorphism)") +
  ylab("log(Canine Dimorphism)") + 
  geom_point(na.rm = TRUE) + 
  facet_wrap(~Family, ncol = 4) +
  theme(legend.position = "none") +
  geom_smooth(method = "lm", fullrange = FALSE, se=FALSE, na.rm = TRUE)
p
```

Particularly within Cebidae and Cercopithecidae, it does appear that there could be a trend for species that exhibit greater body size dimorphism to also exhibit greater canine dimorphism.

**7. Here, I'm creating a new variable named diet_strategy.**
```{r}
d <- d|> mutate(diet_strategy = case_when( #If...
  Fruit > 50.0  ~ "Frugivore", #...fruit makes up >50% of diet, then designate species as "frugivore"
  Leaves > 50.0 ~ "Folivore", #...leaves make up >50% of diet, then designate species as "folivore"
  TRUE ~ "Omnivore", #...neither case is true, designate all others as omnivores
))
```

Now I'm creating boxplots of group size for species with different dietary strategies
```{r}
p <- ggplot(data = d, aes(x = diet_strategy, y = log(MeanGroupSize))) +
  geom_boxplot(na.rm = TRUE) + 
  theme(axis.text.x = element_text(angle = 90)) + #Specifying axis labels
  ylab("log(Mean Group Size)") + 
  xlab("Diet Strategy") 
p
```

The mean group size for all frugivore taxa is less than the mean group size for all folivore taxa.

**8.** 
```{r}
(a <- mutate(d, Binomial = paste(Genus, Species, sep = ",")) |> #creating a new variable, Binomial, which is a concatenation of Genus and Species
  select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) |> #Trimming the data frame to only include these variables
  group_by(Family) |> #Grouping by family
  summarise(avgBrain_Size_Species_Mean = mean(Brain_Size_Species_Mean, na.rm = TRUE), avgBody_mass_male_mean = mean(Brain_Size_Species_Mean, na.rm = TRUE)) |> #Calculating the average value for Brain_Size_Species_Mean and Body_mass_male_mean by Family
  arrange(avgBrain_Size_Species_Mean)) #Arranging by increasing average brain size
```

